### 2021-11-21 draft.tex
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@fb46df1d360ffce6ec710ba72cc4c221df2f0b2a 🚀
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@22d0e6df93e011819886625b42d04f90849bf725 🚀

The second major challenge in ecological forecasting is that the
underlying dynamics of most ecological processes are unknown and instead
must be inferred from this (sparse) data. Much of the history of
quantitatively modeling ecosystems have been done in the language of
dynamical systems, describing how the value of an observable state of
the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time, yielding
models in the form of differential equations in continuous-time
settings--\(\frac{dx}{dt} = f(x)\)-- or difference equations in
discrete-time settings--\(x_t = f(x_{t-1})\)--where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The initial success of these
forms of models can be traced back to the larger program of ontological
reductionism, which became the default approach to modeling in the
sciences after its early success in physics, which, by the time ecology
was becoming a quantitative science (sometime in the 20th century,
depending on who you ask), became the foundation for early quantitative
models in ecology.

However, we run into many problems when aiming to apply this type of
model to empirical data in ecology. Ecosystems are perhaps the
quintessential example of system that cannot be understood by iterative
reduction of its components into constituent parts---ecological
phenomena are emergent are the product of different mechanisms operating
a different spatial, temporal, and organizational scales (Levin 1992).
Further, the form of this functional response in real systems is
effectively unknown, and some forms are inherently more ``forecastable''
than others (Beckage \emph{et al.} 2011; Chen \emph{et al.} 2019;
Pennekamp \emph{et al.} 2019). Further this analytical approach to
modeling explicitly ignores known realities: ecological dynamics not
deterministic, many analytic models in ecology assume long-run
equilibrium. Finally, perhaps the biggest challenge in using these
models to describe ecological processes is ecosystems vary across more
variables than the tools of analytic models are suited for. As the
number of variables in an analytic model increases, so does the ability
of the scientist to discern clear relationships between them given a
fixed amount of data, the so-called ``curse'' of dimensionality.

@@

\hypertarget{methods-1}{%
\subsection{Methods}\label{methods-1}}

>>>>

The second major challenge in ecological forecasting is that the
underlying dynamics of most ecological processes are unknown and instead
must be inferred from this (sparse) data. Much of the history of
quantitatively modeling ecosystems have been done in the language of
dynamical systems, describing how the value of an observable state of
the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time, yielding
models in the form of differential equations in continuous-time
settings, \(\frac{dx}{dt} = f(x)\), or difference equations in
discrete-time settings, \(x_t = f(x_{t-1})\), where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The form of this functional
response in real systems is effectively unknown, and some forms are
inherently more ``forecastable'' than others (Beckage \emph{et al.}
2011; Chen \emph{et al.} 2019; Pennekamp \emph{et al.} 2019). The
initial success of these forms of models can be traced back to the
larger program of ontological reductionism, which became the default
approach to modeling in the sciences after its early success in physics,
which, by the time ecology was becoming a quantitative science (sometime
in the 20th century, depending on who you ask), became the foundation
for mathematical models in ecology.

However, we run into many problems when aiming to apply this type of
model to empirical ecological data. Ecosystems are perhaps the
quintessential example of system that cannot be understood by iterative
reduction of its components into constituent parts---ecological
phenomena are emergent: the product of different mechanisms operating a
different spatial, temporal, and organizational scales (Levin 1992).
Further this analytical approach to modeling explicitly ignores known
realities: ecological dynamics not deterministic and many analytic
models in ecology assume long-run equilibrium. Finally, perhaps the
biggest challenge in using these models to describe ecological processes
is ecosystems consist of more dimensions than the tools of analytic
models are suited for. As the number of variables in an analytic model
increases, so does the ability of the scientist to discern clear
relationships between them given a fixed amount of data, the so-called
``curse'' of dimensionality.

@@
